Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data

### 2021-07-05 Manuscript/SS_GAM.tex
Finish revising method section.

We introduce a novel prior for GAMs, particularly for high-dimensional
nonlinear effect modeling with bi-level selection. The proposed prior
extends from the spike-and-slab lasso prior described in Equation
\ref{eq:ssl}. Given the reparameterized deign matrix
\(\boldsymbol{X}_j = \begin{bmatrix} \boldsymbol{X}_j^0 : \boldsymbol{X}_j^\text{pen}\end{bmatrix}^T\)
for the \(j\)th variable, we impose a two-part SSL prior to the
coefficients \(\boldsymbol{\beta}_j\). Specifically, we impose
independent group priors on the null space coefficients and on the
penalized space coefficients respectively, \[
\begin{aligned}
  \beta^0_{j} |\gamma^0_{j},s_0,s_1 &\sim DE(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1)\\
  \beta^\text{pen}_{jk} | \gamma^\text{pen}_{j},s_0,s_1 &\sim DE(0,(1-\gamma^\text{pen}_{j}) s_0 + \gamma^\text{pen}_{j} s_1), 
\end{aligned}
\] where \(\gamma^0_{j}\in\{0,1\}\) and
\(\gamma^\text{pen}_{j}\in \{0,1\}^{K_j}\) are latent indicator
variables, indicating if the model includes the linear effect and the
non-linear effect of the \(j\)th variable respectively. ; \(s_0\) and
\(s_1\) are scale parameters, assuming \(0 < s_0 < s_1\) and given.
These scale parameters \(s_0\) and \(s_1\) can be treated as tuning
parameters and optimized via cross-validation. A discussion of how to
choose the scale parameters comes in Section
\ref{selecting-optimal-scale-values}. To note, this prior differs from
previous group spike-and-slab lasso priors{[}TODO: add Dr.~Yi's
citation{]}, where the \(\boldsymbol{\beta}^0_j\) and
\(\boldsymbol{\beta}^\text{pen}_j\) shares the same indicator variables
\(\gamma_j^0\), \(\gamma_j^\text{pen}\) respectively. It is possible to
relax the assumption such that each coefficient \(\beta\) to have its
own latent indicator \(\gamma\), but at the cost ofcomplicating the
bi-level function selection. Conversely, it is possible to add a more
restrictive assumption on the priors, assuming that one indicator
variable decides the inclusion of both the linear effect and non-linear
effect, i.e.~\(\gamma_j = \gamma^0_j = \gamma^{pen}_j\). This
converges\footnote{Is ``becomes'' better?} to the SB-GAM
\citep{Bai2021}.

@@

The rest of the hierchical prior follows the traditional spike-and-slab
lasso piror: we set up the hyper-prior of \(\boldsymbol{\gamma}\), to
allow local adaption of the shrinkage, using a bernoulli distribution.
The two indicators of the \(j\)th predictor, \(\gamma^{0}_j\) and
\(\gamma^{pen}_j\), shares the same probability parameter, \[
\begin{aligned}
\gamma_{j}^{0} | \theta_j &\sim Bin(\gamma^{0}_{j}|1, \theta_j)\\
\gamma_{j}^{pen} | \theta_j &\sim Bin(\gamma^{pen}_{j}|1, \theta_j).
\end{aligned}
\]

This is to leverage the fact that the probability of selecting the bases
of a smooth function should be similar, while allowing different penalty
on the null space and penalty space of the spline function. The hyper
prior of \(\gamma_{j}^{0}\) decides the sparsity of the model at the
function selection level, while that of \(\gamma_{j}^{pen}\) decides the
smoothness of the spline function at basis function level. Meanwhile, we
specify that \(\gamma_{j}^0\) and \(\gamma_{j}^{pen}\) are independently
distributed for analytic simplicity. We further specify the parameter
\(\theta_j\) follows a beta distribution with given shape parameters
\(a\) and \(b\), \[
\theta_j \sim Beta(a, b).
\] The beta distribution is a conjugate prior for the binomial
distribution and hence provides some computation convenience. For
simplicity, we focus on a special case of beta distribution, uniform
(0,1). When the variable have large effects in any of the bases, the
parameter \(\theta_j\) will be estimated large, which in turn encourages
the model to include the rest of bases, achieving the local adaption
among spline bases. Hereafter, we refer Bayesian hierarchical
generalized additive models with the spike-and-slab lasso prior as the
BHAM-SSL.

\newpage

@@

Considering the dependency of spline bases, the grouping nature of the
bases should be taken care of. Naturally, the Bayesian hierarchical
models allow us easily to take account of the dependency of the spline
bases as grouping information and provide more reliable estimation. Our
hierarchical GAMs specify the spike-and-slab mixture normal prior on the
coefficients: for the coefficients of the \(j\)th predictor,
\(\beta^0_j\) and \(\beta^{pen}_{jk}\), we have \[
\begin{aligned}
  \beta^0_{j} |\gamma^0_{j},s_0,s_1 &\sim N(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1)\\
  \beta^{pen}_{jk} | \gamma^{pen}_{j},s_0,s_1 &\sim N(0,(1-\gamma^{pen}_{j}) s_0 + \gamma^{pen}_{j} s_1), 
\end{aligned}
\] where \(\gamma^0_{j}\in\{0,1\}\) and
\(\gamma^{pen}_{j}\in \{0,1\}^{K_j}\) are latent indicator variables,
indicating if the model includes the linear effect and non-linear effect
of the \(j\)th variable respectively; \(s_0\) and \(s_1\) are scale
parameters, assuming \(0 < s_0 < s_1\) and given. Thus, the priors are
formulated as a mixture of the shrinkage prior normal(0, \(s_0\)) and
the weakly informative prior normal(0, \(s_1\)), which are spike and
slab components respectively. Like any other spike and slab priors, the
spike density is to contain the minimum to zero effects, while the slab
density is to allow large effects. The scale parameters \(s_0\) and
\(s_1\) can be treated as tuning parameters, which can be optimized via
cross-validation. A discussion of how to choose the scale parameters
comes later.

To note here, unlike the original spike-and-slab prior where each
predictor have only indicator variable, \(\gamma_j\), we have two
indicators associated with each predictor, \(\gamma^0_j\) and
\(\gamma^{pen}_j\) to decide the inclusion of the linear effect and
non-linear effect of the predictor respectively. It is possible to add
more constraints on the model, assuming that one indicator variable
decides the inclusion of both the linear effect and non-linear effect,
i.e.~\(\gamma_j = \gamma^0_j = \gamma^{pen}_j\). Such formulations,
commonly referred as ``all-in-all-out'', has been implemented in many
high dimensional Bayesian GAMs, see Yang \& Narisetty (2020) and Bai
(Work in progress). The ``all-in-all-out'' set-up was criticized for its
inability to answering an important research question, if the effects
are non-linear or not. On the other hand, the model can be much more
flexible by allowing each dimension of the penalized space to have an
independent indicator, i.e.~\(\gamma^{pen}_{jk}\) of
\(\beta^{pen}_{jk}\) for \(k = 1\dots, K_j\). This prior set-up
generalized to the group spike-and-slab. The additional flexibility
doesn't not necessarily grand the model neither theoretical
justification nor analytic simplicity. The spline penalty assumes a
common smooth shrinkage \(\lambda_j\) applied to all coefficients of the
penalized space, which contradicts to the model setup. Besides, having
multiple indicator variables complicates the decision of if non-linear
effects are necessary.

We set up the hyper-prior of \(\boldsymbol{\gamma}\) to allow local
adaption of the shrinkage, using a binomial distribution. The two
indicators of the \(j\)th predictor, \(\gamma^{0}_j\) and
\(\gamma^{pen}_j\), shares the same probability parameter, \[
\begin{aligned}
\gamma_{j}^{0} | \theta_j &\sim Bin(\gamma^{0}_{j}|1, \theta_j)\\
\gamma_{j}^{pen} | \theta_j &\sim Bin(\gamma^{pen}_{j}|1, \theta_j).
\end{aligned}
\]

This is to leverage the fact that the probability of selecting the bases
of a smooth function should be similar, while allowing different penalty
on the null space and penalty space of the spline function. The hyper
prior of \(\gamma_{j}^{0}\) decides the sparsity of the model at the
function selection level, while that of \(\gamma_{j}^{pen}\) decides the
smoothness of the spline function at individual predictor level.
Meanwhile, we specify that \(\gamma_{j}^0\) and \(\gamma_{j}^{pen}\) are
independently distributed for analytic simplicity. However, this set-up
could possibility result in a situation that is not theoretically
possible: the non-linear component is selected, but the linear component
is not. This can be addressed analytically by forcing to include the
linear component when non-linear component is included. Another possible
solution is to impose an dependent structure of \(\gamma_{j}^{pen}\) on
\(\gamma_{j^{0}}\), i.e.~\(\gamma_j^{pen}|\gamma_{j}^{0}, \theta_j\).

We further specify the parameter \(\theta_j\) follows a beta
distribution with given shape parameters \(a\) and \(b\), \[
\theta_j \sim Beta(a, b).
\] The beta distribution is a conjugate prior for the binomial
distribution and hence provides some computation convenience.
Specifically, we focus on a special case of beta distribution, uniform
(0,1) for simplicity and convenience. To note, when the variable have
large effects in any of the bases, the parameter \(\theta_j\) will be
estimated large, which in turn encourages the model to include the rest
of bases, achieving the local adaption among spline bases. Hereafter, we
refer the hierarchical spline GAM with the spike-and-slab mixture normal
prior as the ssGAM-MN.

One of the critics received by the spike-and-slab mixture normal prior
is that the tails of a normal distribution diminishes to zero too fast,
which renders problem when estimating the large effects. In contrast,
distributions with heavier tails are preferred, for example \(t\)
distribution with small degree of freedom and double exponential
distribution. Specifically, double exponential prior is equivalent to
the LASSO penalty in the frequentist models, providing a sparse
solution.

@@

To alleviate the computation burden, many methodology development
studies focus on the maximum a posteriori (MAP) estimate
\citep{Tipping2004}. EM-based algorithms are prevalent used to speed the
computation. \citet{Rockova2014a} proposed an EM-based algorithm for a
spike-and-slab Gaussian mixture prior, EMVS. The coefficient \(\beta_j\)
independently follows a mixture normal distribution
\(N(0, \sigma^2((1-\gamma_i)v_0 + \gamma_i v_1)\) for
\(0 \leq v_0 < v_1\). The parameter \(\sigma^2\) follows a inverse gamma
prior, \(IG(v/2, v\lambda /2)\). Like in the spike-and slab family, the
latent binary variables \(\boldsymbol{\gamma}\) follows a Bernoulli
prior with a hyper prior \(\theta\) follows a beta distribution, or
simpler the uniform (0,1). In the E step, the latent variable are
treated as the missing data, and calculate the posterior mean of
\(\gamma\); in the M step, a ridge estimator was used to update the
coefficients, and \(\sigma^2, \theta\) are updated accordingly. As the
ridge regression creates small but nonzero coefficients, a further step
is needed to select the variable: EMVS further suggests to threshold the
inclusion probability for variable selection or generate regularization
plot with different values of the spike scale parameter. EMVS is also
compatible with structured prior when dealing with grouped variables.
The group structure can be imposed on the prior of the latent binary
variable \(\gamma\) via either logistic regression or markov random
field.

The same authors \citep{Rockova2018b, Rockova2018} further extended the
continuous spike-and-slab Gaussian prior to the spike-and-slab double
exponential distribution, as known as spike-and-slab lasso (SSL). The
authors also lay the theoretical discussion of comparing spike-and-slab
from the penalty perspective and the Bayesian perspective. The SSL also
mitigated the problem of EMVS that the weak signals are not shrink to
zero. Yi and his group incorporated the SSL prior in the EM-IWLS
algorithm and proposed another EM algorithm, EM-cyclic coordinate
descent, based on the Lasso penalty SSL exhibit. Such application were
used in generalized linear models\citep{Tang2017a}, cox models
\citep{Tang2017} and their grouped
counterparts\citep{Tang2018, Tang2019}. Both EM based algorithm provide
deterministic solutions, which becomes a more popular property as
reproducible research becomes more valued. However, while the EMVS are
not able to provide inference of the coefficients easily, the EM-IWLS
can easily extract the variance-covariance matrix due to the IWLS
algorithm. \citet{Bai2020} provides an up-to-date summary of SSL methods
and theoretical discussion.

Given the proposed model specification, the estimation of the parameters
can be easily achieved with MCMC algorithm. Nevertheless, considering
the computational burden MCMC algorithms bear especially in
high-dimensional analysis, we propose parsimonious solutions, stemming
from previous EM-IWLS \citep{Rockova2014a}and EM-coordinate descent
algorithms\citep{Tang2018, Tang2019}. The computational friendliness
allows for fitting a model with different sets of initial values,
mimicking the posterior sampling in the MCMC algorithms. We are hopeful
that this could improve the sacrificed robustness of the estimation
process due to the MAP estimation used by the EM algorithms in
comparison to the model averaging approach.

Parsimonious computation is always encouraged in high-dimension data
analysis. Normally, Bayesian methodology loses its advantages over
Frequentest penalized model because of the computation cost. Previous
Bayesian spline models heavily relies on the MCMC algorithm to establish
posterior distribution of parameters. One of the exception is Bai (in
progress), who a grouped spike-and-slab LASSO prior \citep{Rockova2018}
to address the grouping structure of spline bases. He focused on the MAP
estimator of the parameters, but failed to address the uncertainty
measure of the estimates. The main difference between Bai's work and the
proposed models are two-folded: 1) the proposed models are feasible to
answer the question if the non-linear effects are necessary for the
predictors, while Bai's work can't due to the ``all-in-all-out''
approach; 2) the proposed models can provide uncertainly measures in
addition to the point estimates of the coefficients, while Bai's work
can't not due to the model fitting algorithm.

In the follow section, we introduced two expectation-maximization (EM)
based algorithms for fast computing, EM-IRLS and EM-coordinate Descent
algorithm. Instead of simulating the posterior distribution via MCMC, we
iteratively estimate the MAP. Compared to the EM-IRLS algorithm, the
EM-coordinate descent provides an expedited model fitting process at the
cost of model uncertainty.

@@

For ssGAM, we define the parameters of interest as
\(\Theta = {\boldsymbol{\beta}, \boldsymbol{\theta}, \phi}\) and
consider the latent binary indicators \(\boldsymbol{\gamma}\) as
nuisance parameters of the model. Our objective is to find the
parameters \(\Theta\) that maximize the posterior density function,or
equivalently, the logarithm of the density function, \[
\begin{aligned}
& \text{argmax}_{\Theta}
\log p(\Theta, \boldsymbol{\gamma}| \textbf{y}, \textbf{X}) \\
&= \log p(\textbf{y}|\boldsymbol{\beta}, \phi) + \log p(\phi) + \sum\limits_{j=1}^p\left[\log p(\beta^0_j|\gamma^0_j)+\sum\limits_{k=1}^{K_j} \log p(\beta^{pen}_{jk}|\gamma^{pen}_{jk})\right]\\
& +\sum\limits_{j=1}^{p} \left[ (\gamma^0_j+\gamma_{j}^{pen})\log \theta_j + (2-\gamma^0_j-\gamma_{j}^{pen}) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log p(\theta_j)
\end{aligned}
\]\\

We use the EM algorithm to find the MAP estimate of \(\Theta\). Since
the latent binary indicators \(\boldsymbol{\gamma}\) are not of our
primary interest, we treat them as as the ``missing data'' in the EM
algorithms. Naturally In the E-step, we calculate the expectation of
posterior density function of
\(\log p(\Theta, \boldsymbol{\gamma}| \textbf{y}, \textbf{X})\) with
respect to the latent indicators \(\boldsymbol{\gamma}\) conditioning on
the values from previous iteration \(\Theta^{t-1}\), \[
E_{\boldsymbol{\gamma}|\Theta^{t-1}}\log p(\Theta, \boldsymbol{\gamma}| \textbf{y}, \textbf{X}) .
\] Hereafter, we use the short hand notation
\(E(\cdot)\equiv E_{\boldsymbol{\gamma}|\Theta^{t-1}}(\cdot)\). In the
M-step, we find the \(\Theta^{t}\) that maximize
\(E_{\boldsymbol{\gamma}|\Theta^{t-1}}\log p(\Theta, \boldsymbol{\gamma}| \textbf{y}, \textbf{X})\).
To note here, the log-posterior of ssGAM (up to additive constants) can
be written as a two-part equation

\[ \log p(\Theta, \boldsymbol{\gamma}| \textbf{y}, \textbf{X}) = Q_1(\beta, \phi) + Q_2 (\gamma,\theta),\]
Where
\[ Q_1(\boldsymbol{\beta}, \phi) = \log p(\textbf{y}|\boldsymbol{\beta}, \phi) + \log p(\phi) + \sum\limits_{j=1}^p\left[\log p(\beta^0_j|\gamma^0_j)+\sum\limits_{k=1}^{K_j} \log p(\beta^{pen}_{jk}|\gamma^{pen}_{jk})\right]\]
and \[
Q_2(\boldsymbol{\gamma},\boldsymbol{\theta}) = \sum\limits_{j=1}^{p} \left[ (\gamma^0_j+\gamma_{j}^{pen})\log \theta_j + (2-\gamma^0_j-\gamma_{j}^{pen}) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log p(\theta_j) .\]
\(Q_1\) and \(Q_2\) are respectively, the log-likelihood of model
coefficients and the dispersion parameter \(\boldsymbol{\beta}, \phi\),
and latent indicators \(\boldsymbol{\gamma}, \theta\). Conditioning on
\(\gamma_{jk}\), \(Q_1\) and \(Q_2\) are independent and can be
maximized separately for \(\boldsymbol{\beta}, \phi\) and
\(\boldsymbol{\theta}\).

The E- and M- steps are iterated until the algorithm converge. Depending
on the choice of the mixture priors, either mixture normal or mixture
double exponential, the maximization in the M-step can take different
algorithms, IRLS for uncertainty measures and coordinate descent for
faster computation.

\hfill\break

The main idea of the EM-IRLS algorithm is to use the iterative
re-weighted least square algorithm to find the estimate
\(\boldsymbol{\beta}, \phi\) that maximizes \(E(Q_1)\). Recall the prior
set-up of mixture normal prior, we have the prior densities of
\(\boldsymbol{\beta}\) conditioning on \(\boldsymbol{\gamma}\)

\[
\begin{aligned}
p(\boldsymbol{\beta }| \boldsymbol{\gamma}) &\propto \prod\limits_{j=1}^{p}\left[{S^0_j}^{-1/2}\exp(-1/2({\beta^0_{j}}^2/S^0_{j}))\prod\limits_{k=1}^{K_j}S_{j}^{-1/2}\exp(-1/2(\beta_{jk}^2/S_{j}))\right],%\\
% p(\bs\gamma | \bs \theta) & = \prod\limits_{j=1}^{p}\prod\limits_{k=1}^{K_j} \theta_j^{\gamma_{jk}} (1-\theta_j)^{1-\gamma_{jk}}.
\end{aligned}
\] and \begin{align*}
E(Q_1) &= \log p(\textbf{y}|\boldsymbol{\beta}, \phi) + \log p(\phi) + \sum\limits_{j=1}^{p} \left[E({S^0_j}^{-1}){\beta^0_j}^2+\sum\limits_{k=1}^{K_j}E(S^{-1}_{j})\beta_{jk}^2\right]\\
E(Q_2) &= \sum\limits_{j=1}^{p}\sum\limits_{k=1}^{K_j} \left[ E(\gamma_{jk})\log \theta_j + (1-E(\gamma_{jk})) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log p(\theta_j)
\end{align*} where
\(S_{j}^0 = (1-\gamma^{0}_{j}) s_0 + \gamma^{0}_{j} s_1\) and
\(S_{j} = (1-\gamma^{pen}_{j}) s_0 + \gamma^{pen}_{j} s_1\). To
calculate two unknown quantities \(E({S_j^0}^{-1})\) and
\(E(S^{-1}_j)\), the posterior probability
\(p^0_{j} \equiv p(\gamma^{0}_{j}=1|\Theta^{t-1})\) and
\(p_{j} \equiv p(\gamma^{pen}_{j}=1|\Theta^{t-1})\) are necessary, which
can be derived via Bayes' theorem. The calculation for \(p_j\) is
slightly different from that of \(p^0_j\), as \(p_j\) depends on the
value of \(\beta_{jk}\) for \(k=1, \dots, K_j\) and \(p^0_j\) only
depends on \(\beta_j^0\). The calculation follows the equations below.

and
\(Pr(\gamma_{j}^{0} = 1|\theta_j) =Pr(\gamma_{j}^{pen} = 1|\theta_j) = \theta_j\),
\(Pr(\gamma_{j}^{0} = 0|\theta_j) = Pr(\gamma_{j}^{pen} = 0|\theta_j) = 1-\theta_j\),
\(f(\beta|\gamma=1, s_1) = \text{Normal}(\beta|0 , s_1)\),
\(f(\beta|\gamma=0, s_0) = \text{Normal}(\beta|0 , s_0)\). It is trivial
to show \[
\begin{aligned}
&E(\gamma^0_{j})  = p^0_{j} & &E(\gamma^{pen}_{j})  = p_{j}\\
&E({S^0}^{-1}_{j}) = \frac{1-p^{0}_{j}}{s_0} + \frac{p^{0}_{j}}{s_1} & &E(S^{-1}_{j}) = \frac{1-p_{j}}{s_0} + \frac{p_{j}}{s_1}.
\end{aligned}
\]

>>>>

We introduce a novel prior for GAMs, particularly for high-dimensional
nonlinear effect modeling with bi-level selection. The proposed prior
extends from the spike-and-slab lasso prior described in Equation
\ref{eq:ssl}. Given the reparameterized deign matrix
\(\boldsymbol{X}_j = \begin{bmatrix} \boldsymbol{X}_j^0 : \boldsymbol{X}_j^\text{pen}\end{bmatrix}^T\)
for the \(j\)th variable, we impose a two-part SSL prior to the
coefficients \(\boldsymbol{\beta}_j\). Specifically, we impose
independent group priors on the null space coefficients and on the
penalized space coefficients respectively,

\begin{align}\label{eq:bham_ssl}
  \beta^0_{j} |\gamma^0_{j},s_0,s_1 &\sim DE(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1)\\
  \beta^\text{pen}_{jk} | \gamma^\text{pen}_{j},s_0,s_1 &\sim DE(0,(1-\gamma^\text{pen}_{j}) s_0 + \gamma^\text{pen}_{j} s_1), 
\end{align}

where \(\gamma^0_{j}\in\{0,1\}\) and
\(\gamma^\text{pen}_{j}\in \{0,1\}^{K_j}\) are latent indicator
variables, indicating if the model includes the linear effect and the
non-linear effect of the \(j\)th variable respectively. ; \(s_0\) and
\(s_1\) are scale parameters, assuming \(0 < s_0 < s_1\) and given.
These scale parameters \(s_0\) and \(s_1\) can be treated as tuning
parameters and optimized via cross-validation. A discussion of how to
choose the scale parameters comes in Section
\ref{selecting-optimal-scale-values}. To note, this prior differs from
previous group spike-and-slab lasso priors{[}TODO: add Dr.~Yi's
citation{]}, where the \(\boldsymbol{\beta}^0_j\) and
\(\boldsymbol{\beta}^\text{pen}_j\) shares the same indicator variables
\(\gamma_j^0\), \(\gamma_j^\text{pen}\) respectively. It is possible to
relax the assumption such that each coefficient \(\beta\) to have its
own latent indicator \(\gamma\), but at the cost ofcomplicating the
bi-level function selection. Conversely, it is possible to add a more
restrictive assumption on the priors, assuming that one indicator
variable decides the inclusion of both the linear effect and non-linear
effect, i.e.~\(\gamma_j = \gamma^0_j = \gamma^\text{pen}_j\). This
converges\footnote{Is ``becomes'' better?} to the SB-GAM
\citep{Bai2021}.

@@

The rest of the hierchical prior follows the traditional spike-and-slab
lasso piror: we set up the hyper-prior of \(\boldsymbol{\gamma}\), to
allow local adaption of the shrinkage, using a bernoulli distribution.
The two indicators of the \(j\)th predictor, \(\gamma^{0}_j\) and
\(\gamma^\text{pen}_j\), shares the same probability parameter, \[
\begin{aligned}
\gamma_{j}^{0} | \theta_j &\sim Bin(\gamma^{0}_{j}|1, \theta_j)\\
\gamma_{j}^\text{pen}| \theta_j &\sim Bin(\gamma^\text{pen}_{j}|1, \theta_j).
\end{aligned}
\]

This is to leverage the fact that the probability of selecting the bases
of a smooth function should be similar, while allowing different penalty
on the null space and penalty space of the spline function. The hyper
prior of \(\gamma_{j}^{0}\) decides the sparsity of the model at the
function selection level, while that of \(\gamma_{j}^\text{pen}\)
decides the smoothness of the spline function at basis function level.
Meanwhile, we specify that \(\gamma_{j}^0\) and
\(\gamma_{j}^\text{pen}\) are independently distributed for analytic
simplicity. We further specify the parameter \(\theta_j\) follows a beta
distribution with given shape parameters \(a\) and \(b\), \[
\theta_j \sim Beta(a, b).
\] The beta distribution is a conjugate prior for the binomial
distribution and hence provides some computation convenience. For
simplicity, we focus on a special case of beta distribution, uniform
(0,1). When the variable have large effects in any of the bases, the
parameter \(\theta_j\) will be estimated large, which in turn encourages
the model to include the rest of bases, achieving the local adaption
among spline bases. Hereafter, we refer Bayesian hierarchical
generalized additive models with the spike-and-slab lasso prior as the
BHAM-SSL.

@@

With the re-parameterization step of the basis function matrix \(X\), it
is possible to generalized the SSL prior to other priors, for example
normal priors for ridge-type regularizations and mixture normal prior
for spike-and-slab regularization. These priors would work better in low
and medium dimensional settings where the sparse assumption is not
necessary. Here we focus on the mixture normal prior as a demonstration
of applying continous spike-and-slab prior in additive models.
\footnote{TODO: enter a sentence introducing EMVS and spike-and-slab
  mixture normal prior}.

A spike-and-slab mixture normal prior in additive models can be
expressed as \[
\begin{aligned}
  \beta^0_{j} |\gamma^0_{j},s_0,s_1 &\sim N(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1)\nonumber\\
  \beta^\text{pen}_{jk} | \gamma^\text{pen}_{j},s_0,s_1 &\sim N(0,(1-\gamma^\text{pen}_{j}) s_0 + \gamma^\text{pen}_{j} s_1).
\end{aligned}
\] Similar to the BHAM-SSL prior in Equation \ref{eq:bham_ssl},
\(0 < s_0 < s_1\) are tuning parameters and can be optimized via
cross-validation. One of the critics received by the spike-and-slab
mixture normal prior is that the tails of a normal distribution
diminishes to zero too fast, which renders problem when estimating the
large effects. Distributions with heavier tails can be used as an
alternative, for example mixture \(t\) distribution with small degree of
freedom.

@@

Given the proposed model specifications, these models can be easily
fitted with MCMC algorithms. Nevertheless,the computational burden MCMC
algorithms bear creates scalability issues. George and
McCulloch\citep{George1997} examined the computation speed for various
MCMC algorithms with continous spike-and-slab priors, and suggested MCMC
algorithms works well for medium size of predictors (\(p\)=25). However,
it is not feasible to high-dimensionality data analysis where the number
of predictors easily exceeds 100. Sepcific to additive models, each
predictor would expand to mutliple new ``predictors'' via basis
functions, creating greater computational demands. \citet{Scheipl2013}
demonstrated the computational demands of MCMC algorithm for fitting
spike-and-slab additive models grow exponentially as \(p\) increases
modestly via monte carlo simulation studies. Hence, we feel compelled to
develop scalable and parsimonious algorithms for fitting Bayesian
hierarchical additive models in high-dimensional settings.

Alternative to the sampling algorithms for fitting Bayesian models,
optimization algorithms focus on the maximum a posteriori (MAP)
estimates and speed up the process at the cost of uncertainty measures.
The earliest work for fitting spike-and-slab models using optimization
algorithm is EMVS \citep{Rockova2014a}. \citet{Rockova2014a} proposed an
expectation-maximization (EM) based algorithm to fit models that use
continous mixture normal priors. In the E step, the latent variables
\(\gamma\) are treated as the missing data, and calculate the posterior
mean of the density function; in the M step, a ridge estimator was used
to update the coefficients, and \(\sigma^2, \theta\) are updated
accordingly. The same authors \citep{Rockova2018b, Rockova2018} further
combined the EM algorithm with coodirnate descent (CD) algoirthm for
fitting SSL models. Yi and his group incorporated the SSL prior in the
EM-Iterative Weighted Least Square(IWLS) algorithm and proposed another
EM algorithm, EM-cyclic coordinate descent, based on the Lasso penalty
SSL exhibit. Such applications were used in generalized linear
models\citep{Tang2017a}, cox models \citep{Tang2017} and their grouped
counterparts\citep{Tang2018, Tang2019}. Both EM based algorithms provide
deterministic solutions, which becomes a more popular property as
reproducible results. \citet{Bai2020} provides an up-to-date summary of
SSL methods and theoretical discussion.

In the follow section, we expand the two EM-based algorithms, EM-CD and
EM-IWLS algorithms, to fit BHAMs. To note, the two proposed algorithm
provides different utilities. The EM-CD algorithm is specifically for
fitting BHAM-SSL with an expediated performance, recommending to use in
high and ultra-high dimensional setting. SB-GAM\citep{Bai2020, Bai2021}
also used EM-CD algorithm. The main difference between proposed EM-CD
algorithm and SB-GAM is that SB-GAM uses a blocked CD algorithm for
their group prior. In comparison, our prior is pairwise independent
requiring no speical treatment. A sepcifica concern of EM-CD algorithm
is that it provides no information for inference. In the contrast, the
EM-IWLS provides a complement to the problem. The EM-IWLS can easily
extract the variance-covariance matrix due to the IWLS algorithm, and
used to construct Wald type inference. Moreover, the EM-IWLS is a more
general model fitting algorithm that can be used for fitting not only
SSL and continuous SS pirors but also bayesian ridge and lasso priors.

@@

For BHAMs, we define the parameters of interest as
\(\Theta = {\boldsymbol{\beta}, \boldsymbol{\theta}, \phi}\) and
consider the latent binary indicators \(\boldsymbol{\gamma}\) as
nuisance parameters of the model. Our objective is to find the
parameters \(\Theta\) that maximize the posterior density function, or
equivalently, the logarithm of the density function, \[
\begin{aligned}
& \text{argmax}_{\Theta}
\log p(\Theta, \boldsymbol{\gamma}| \textbf{y}, \textbf{X}) \\
&= \log p(\textbf{y}|\boldsymbol{\beta}, \phi) + \log p(\phi) + \sum\limits_{j=1}^p\left[\log p(\beta^0_j|\gamma^0_j)+\sum\limits_{k=1}^{K_j} \log p(\beta^{pen}_{jk}|\gamma^{pen}_{jk})\right]\\
& +\sum\limits_{j=1}^{p} \left[ (\gamma^0_j+\gamma_{j}^{pen})\log \theta_j + (2-\gamma^0_j-\gamma_{j}^{pen}) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log p(\theta_j)
\end{aligned}
\]\\
For simplicity, we consider the null space to be 1-dimension.

We use the EM algorithm to find the MAP estimate of \(\Theta\). Since
the latent binary indicators \(\boldsymbol{\gamma}\) are not of our
primary interest, we treat them as the ``missing data'' in the EM
algorithms. Naturally In the E-step, we calculate the expectation of
posterior density function of
\(\log p(\Theta, \boldsymbol{\gamma}| \textbf{y}, \textbf{X})\) with
respect to the latent indicators \(\boldsymbol{\gamma}\) conditioning on
the values from previous iteration \(\Theta^{t-1}\), \[
E_{\boldsymbol{\gamma}|\Theta^{t-1}}\log p(\Theta, \boldsymbol{\gamma}| \textbf{y}, \textbf{X}) .
\] Hereafter, we use the short hand notation
\(E(\cdot)\equiv E_{\boldsymbol{\gamma}|\Theta^{t-1}}(\cdot)\). In the
M-step, we find the \(\Theta^{t}\) that maximize
\(E\log p(\Theta, \boldsymbol{\gamma}| \textbf{y}, \textbf{X})\). To
note here, the log-posterior density of BHAMs (up to additive constants)
can be written as a two-part equation

\[ \log p(\Theta, \boldsymbol{\gamma}| \textbf{y}, \textbf{X}) = Q_1(\beta, \phi) + Q_2 (\gamma,\theta),\]
Where
\[ Q_1(\boldsymbol{\beta}, \phi) = \log p(\textbf{y}|\boldsymbol{\beta}, \phi) + \log p(\phi) + \sum\limits_{j=1}^p\left[\log p(\beta^0_j|\gamma^0_j)+\sum\limits_{k=1}^{K_j} \log p(\beta^{pen}_{jk}|\gamma^{pen}_{jk})\right]\]
and \[
Q_2(\boldsymbol{\gamma},\boldsymbol{\theta}) = \sum\limits_{j=1}^{p} \left[ (\gamma^0_j+\gamma_{j}^{pen})\log \theta_j + (2-\gamma^0_j-\gamma_{j}^{pen}) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log p(\theta_j) .\]
\(Q_1\) and \(Q_2\) are, respectively, the log-likelihoods of model
coefficients and the dispersion parameter \(\boldsymbol{\beta}, \phi\),
and latent indicators \(\boldsymbol{\gamma}, \theta\). Conditioning on
\(\gamma_{jk}\), \(Q_1\) and \(Q_2\) are independent and can be
maximized separately for \(\boldsymbol{\beta}, \phi\) and
\(\boldsymbol{\theta}\).

The E- and M- steps are iterated until the algorithm converge. Depending
on the choice of the mixture priors, the maximization in the M-step can
take different algorithms, CD for faster computation and IWLS for
uncertainty measures.
